
## Idea / Definition

**Idea:** Rather than giving one value for a parameter, we aim to give two bounds
$B_1$ and $B_2$ and we hope that the true value is between the two


1. Random interval

**Definition:** Let $B_1 = m(Y_1, ..., Y_n)$ et  $B_2 = M(Y_1, ..., Y_n)$ two r.v. We define a random interval for $\theta$ with the couple $(B_1, B_2)$.
We call $P(B_1 < \theta < B_2)$ the level of confidence.

2. Confidence interval

**Definition:** A confidence interval at level $1-\alpha$ for $\theta$ is a realisation
$[b_1, b_2]$ of a random interval with confidence $1-\alpha$

## Confidence interval for the mean knowing the variance (1)

* Data $y_1, ..., y_n$
* Estimator $$\bar{Y} = \sum Y_i /n$$
* We suppose that $V(Y_i) = \sigma^2$ is known
* Model for the estimator [using TCL]

$$\bar{Y} \sim \mathcal{N}(\mu, \sigma^2/\sqrt{n})$$


$$ \frac{\bar{Y} - \mu}{\sigma/\sqrt{n}} \sim\mathcal{N}(0, 1)$$

## Confidence interval for the mean knowing the variance (2)


- Visually

```{r, echo=F}
x <- seq(-4, 4, by=0.01)
plot(x, dnorm(x), type="l", col="blue")
abline(v=qnorm(c(0.025, 0.975)), col="red")
abline(h=0, lty=1, col="black")

```


## Confidence interval for the mean knowing the variance (3)

* So we have
$$ P( u_{\frac{\alpha}{2}} \leq \frac{\bar{Y} - \mu}{\frac{\sigma}{\sqrt{n}}} 
\leq u_{1-\frac{\alpha}{2}}) = 1- \alpha  $$

* If we study the two inequalities

$$ u_{\frac{\alpha}{2}} \leq \frac{\bar{Y} - \mu}{\frac{\sigma}{\sqrt{n}}} \qquad \text{and} \qquad
\frac{\bar{Y} - \mu}{\frac{\sigma}{\sqrt{n}}} \leq u_{1-\frac{\alpha}{2}} $$

* We get

 $$ \mu \leq \bar{Y} - u_{\frac{\alpha}{2}}{\frac{\sigma}{\sqrt{n}}} \qquad \text{and} \qquad
\bar{Y} - u_{1-\frac{\alpha}{2}}{\frac{\sigma}{\sqrt{n}}} \leq \mu$$

## Confidence interval for the mean knowing the variance (4)

* We get
$$ P( \bar{Y} - u_{1-\frac{\alpha}{2}}{\frac{\sigma}{\sqrt{n}}} \leq \mu 
\leq \bar{Y} - u_{\frac{\alpha}{2}}{\frac{\sigma}{\sqrt{n}}}) = 1 - \alpha  $$
$$ B_1 = \bar{Y} - u_{1-\frac{\alpha}{2}}{\frac{\sigma}{\sqrt{n}}} \qquad \text{and} \qquad 
B_2 = \bar{Y} - u_{\frac{\alpha}{2}}{\frac{\sigma}{\sqrt{n}}} $$

* The interval is larger for
	* larger $\sigma$ 
	* smaller $\alpha$ 
	* smaller $n$


## Exercice: Check that a c.i returned by a given approach works reasonably well

## Exercice: First implement in R the previous c.i

Our two bounds are:

$$ B_1 = \bar{Y} - u_{1-\frac{\alpha}{2}}{\frac{\sigma}{\sqrt{n}}} \qquad \text{and} \qquad 
B_2 = \bar{Y} - u_{\frac{\alpha}{2}}{\frac{\sigma}{\sqrt{n}}} $$

```{r}

  y <- rnorm(10) ## 0 mean and sd=1
  n <- length(y)
  theta.h <- mean(y)
  alpha <- 0.05
  b1 <- theta.h + qnorm(p=alpha/2)/sqrt(n)
  b2 <- theta.h + qnorm(p=1-alpha/2)/sqrt(n)
  

```


## Exercice: Check using simulations that our c.i works reasonably well (1)

- What should we check exactly ?
- The claim is that the probability that the random interval with confidence level $1-\alpha$
contains the true mean with probability $1 - \alpha$
- We got our c.i assuming

$$\bar{Y} \sim \mathcal{N}(\mu, \sigma^2/\sqrt{n})$$

## Ex: Check that our c.i works reasonably well (2)

A function to do that simulating uniform $Y_i$:

```{r}
one.simu <- function(n=2, alpha=0.05){
  y <- runif(n, -0.5, 0.5)*sqrt(12) ## 0 mean and sd=1
  theta <- mean(y)
  b1 <- theta + qnorm(p=alpha/2)/sqrt(n)
  b2 <- theta + qnorm(p=1-alpha/2)/sqrt(n)
  return((b1 < 0) & (0 < b2))
}
```

## Ex: Check that our c.i works reasonably well (3)

A few test

```{r}
## n=2, clearly not perfect
res1 <- replicate(10^5, one.simu(2))
mean(res1)

## n=10, farily close
res2 <- replicate(10^5, one.simu(10))
mean(res2)
```



## Ex: Check that our c.i works reasonably well (4)

If we now test for many values of $n$

```{r, cache=TRUE}
InsideProb <- numeric(10)
for(n in c(1:10)){
  InsideProb[n] <- mean( replicate(10^4, one.simu(n)) )
}
```



## Ex: Check that our c.i works reasonably well (5)


```{r}
plot(InsideProb, type="b", lwd=3, col="red", xlab="n", 
     ylim=c(0.8, 1))
abline(h=0.95, lty=2)
```

For large enough (in fact not so large) it works.
With probability 95% the interval

## H-Ex: Check that our c.i works reasonably well (6)

 - We used a uniform distribution for the $Y_i$
 - Test for two times a Bernoulli r.v. of parameter $0.5$
 

```{r}
one.simu <- function(n=2, alpha=0.05){
  y <- 2*rbinom(n=n, size=1, prob=0.5) ## 0 mean and sd=1
  theta <- mean(y)
  b1 <- theta + qnorm(p=alpha/2)/sqrt(n)
  b2 <- theta + qnorm(p=1-alpha/2)/sqrt(n)
  return((b1 < 1) & (1 < b2))
}
```

## H-Ex: Check that our c.i works reasonably well (7)

```{r, cache=TRUE}
InsideProb <- numeric(30)
for(n in c(1:30)){
  InsideProb[n] <- mean( replicate(10^4, one.simu(n)) )
}
```


## H-Ex: Check that our c.i works reasonably well (8)


```{r}
plot(InsideProb, type="b", lwd=3, col="red", xlab="n", 
     ylim=c(0.8, 1))
abline(h=0.95, lty=2)
```

##

## Confidence interval for the mean not knowing the variance 

- If $\sigma$ is not known, similar calculations using the T distribution leads to a confidence interval:

- Namely we start from 
$$ \frac{\bar{X} - \mu}{\sigma/\sqrt{n}} \sim\mathcal{T}_{n-1}$$

## Confidence interval for the mean not knowing the variance 

**In R**

```{r}
x <- runif(10)
t.test(x)$conf.int
```


## Homework exercice and the student confidence interval

## Homework Exercice: Student confidence interval


1. Check that the Student c.i "works" when simulating $Y_i$ as 

- independent Student r.v of degree $2.1$ (in R rt).
- independant $\chi^2$ r.v of degree $3$ (in R rchisq)

2. Study the effet of $n$.

```{r}
one.simu <- function(n=2, alpha=0.05){
  y <- rt(n=n, df=2.1) ## 0 mean and sd=1
  CI <- t.test(y)$conf.int
  return((CI[1] < 0) & (0 < CI[2]))
}
```

## H-Ex: Student c.i

We test for various $n$

```{r, cache=TRUE}
InsideProb <- numeric(30)
InsideProb[1] <- 1
for(n in c(2:30)){
  InsideProb[n] <- mean( replicate(10^4, one.simu(n)) )
}
  
```


## H-Ex: Student c.i

We plot

```{r}
plot(InsideProb, type="b", lwd=3, col="red", xlab="n", 
     ylim=c(0.8, 1))
abline(h=0.95, lty=2)
```

##

## Confidence intervals 

- Many statistical methods provide
confidence intervals
- Computationnal or mathematical derivation of those c.i can be complex
- From an application point of view always the same principle


**Principle**
A statistical model with some assumptions on the signal

1. Check that those assumptions are reasonable for your application
2. In doubt check using simulations that this is working


## Exercice on polymer

## Student interval

* Polymer viscosity

Is the viscocity of the polymer in the interval $[75, 95]$ ?

```{r}
## the data was
y <- c(78, 85, 91, 76)
t.test(y)$conf.int[1:2]
```


What do we conclude ?


##

## An other example simple linear regression

* Linear regression

$Y_i = \alpha x_i + \beta + \varepsilon_i$
$$\varepsilon_i \sim \mathcal{N}(0, \sigma^2) \quad i.i.d$$

```{r, echo=FALSE}
x <- runif(100, 0, 3)
y <- rnorm(100)+2*x
model <- lm(y ~ x)
```

```{r, fig.width=3, fig.height=3, fig.cap="Regression"}
model <- lm(y ~ x) ## Regression
plot(x, y, pch=20); abline(model, col="red"); ## graphe
```

## Simple linear regression (2)


```{r}
confint(model) ## IC à 95%
```

- Can we conclude that the slope is different from 0?
- What about the origin ?



## A schematic view of what a 95% confidence interval does


```{r, echo=FALSE,fig.width=4.5, fig.height=3.5, fig.cap="200 confidence intervals", cache=TRUE}
n <- 1000
IC <- matrix(nrow=n, ncol=2)
for(i in 1:n) IC[i, ] <- t.test(rnorm(5))$conf.int[1:2] ## exp 1
plot(x=c(),ylim=c(-5, 5), xlim=c(-1, 1), ylab="B1 and B2", xlab="Draws of c.i")
x <- seq(-1, 1, length.out=n)
couleur <- c("red", "grey")[(IC[, 1] < 0 & IC[, 2] >0)+ 1]
lwd <- c(2, 1)[(IC[, 1] < 0 & IC[, 2] >0)+ 1]
out.ic <- which(couleur =="red")
arrows(x, IC[, 1], x, IC[, 2], col=couleur, code=3, length=0.05, lwd=lwd);
arrows(x[out.ic], IC[out.ic, 1], x[out.ic], IC[out.ic, 2], col="red", code=3, length=0.05, lwd=2);
abline(h=0, lty=3, lwd=1, col="blue")

```

